Basic regression in Gaussian processes

### 2019-10-26 notebook/gp-manual-vs-greta.Rmd
greta gp exploration

---
layout: post
title: Basic regression in Gaussian processes 
category: ecology
tag: nonparametric-bayes
code: true
output: github_document
---

```{r}
library(tidyverse)
library(greta)
library(greta.gp)

require(MASS)
require(reshape2)
require(ggplot2)
set.seed(12345)
```

@@

Generate (draw) a number of functions from the process

```{r}
values <- mvrnorm(3, rep(0, length=length(x_predict)), COV)
```

@@

fig2a <- ggplot(dat,aes(x=x,y=value)) +
  geom_rect(xmin=-Inf, xmax=Inf, ymin=-2, ymax=2, fill="grey80") +
  geom_line(aes(group=variable))  +
  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
  xlab("input, x") +
  theme_bw()
fig2a
```

### Posterior distribution given the data

@@

## No observation noise

@@

```{r}
values <- mvrnorm(3, Ef, Cf)
```

@@

@@

ggplot(dat,aes(x=x,y=value)) + 
  geom_ribbon(data = gp, 
              aes(x, 
                  y = Ef, 
                  ymin = Ef - sigma, 
                  ymax = Ef + sigma
              ),
              fill="grey80") +
  geom_line(aes(color=variable)) + #REPLICATES
  geom_line(dat = gp, aes(x=x,y=Ef), size=1) + #MEAN
  geom_point(data=obs,aes(x=x,y=y)) +  #OBSERVED DATA
  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
  xlab("input, x")
```

![plot of draws from the posterior, with no proces noise](http://farm9.staticflickr.com/8237/8591045344_918138a53a_o.png)

@@

fig2c <- ggplot(dat,aes(x=x,y=value)) +
  geom_ribbon(data=gp, 
              aes(x=x_predict, y=Ef, ymin=Ef-sigma, ymax=Ef+sigma),
              fill="grey80") + # Var
  geom_line(aes(color=variable)) + #REPLICATES
  geom_line(data=gp,aes(x=x_predict,y=Ef), size=1) + #MEAN
  geom_point(data=obs,aes(x=x,y=y)) +  #OBSERVED DATA
  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
  xlab("input, x")
fig2c
```

Note that unlike the previous case, the posterior no longer collapses completely around the neighborhood of the test points.

>>>>

---
layout: post
title: Basic regression in Gaussian processes 
category: ecology
tag: nonparametric-bayes
code: true
output: html_document
---

First I take a look at doing some simple tasks involving Gaussian processes in base R and sample from them with basic multi-variate normal random draw, `mvrnorm`.  Then, I speculate (unsuccessfully) on how to do this in greta.

```{r setup, message = FALSE, warning = FALSE}

require(MASS)
require(reshape2)
require(ggplot2)

knitr::opts_chunk$set(message=FALSE, warning=FALSE)

set.seed(12345)

```

@@

Generate (draw) a number of functions from the process (note that this is not conditioned on any data, and we are leaving the hyperparameter `l` fixed)

```{r}
values <- mvrnorm(200, rep(0, length=length(x_predict)), COV)
```

@@

```{r}
dat <- data.frame(x=x_predict, t(values)) %>%
  tidyr::pivot_longer(-x, names_to = "rep", values_to = "value") %>% 
  mutate(rep = as.numeric(as.factor(rep)))

ggplot(dat,aes(x=x,y=value)) +
  geom_line(aes(group=rep), color =  rgb(0.7, 0.1, 0.4), alpha = 0.4) 
```

### Posterior distribution conditional on data

@@

No observation noise
--------------------

@@

```{r}
values <- mvrnorm(200, Ef, Cf)
```

@@

```{r}
dat <- data.frame(x=x_predict, t(values)) %>%
  tidyr::pivot_longer(-x, names_to = "rep", values_to = "value") %>% 
  mutate(rep = as.numeric(as.factor(rep)))

@@

ggplot(dat,aes(x=x,y=value)) + 
  #geom_line(aes(group=rep), color =  rgb(0.7, 0.1, 0.4), alpha = 0.2) + #REPLICATES
  geom_ribbon(data = gp, 
              aes(x, 
                  y = Ef, 
                  ymin = Ef - sigma, 
                  ymax = Ef + sigma),
              fill="blue", alpha = 0.1) +
  geom_line(dat = gp, aes(x=x,y=Ef), size=1) + #MEAN
  geom_point(data=obs,aes(x=x,y=y)) +  #OBSERVED DATA
  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
  xlab("input, x")
```

@@

ggplot(dat,aes(x=x,y=value)) +
  geom_ribbon(data=gp, 
              aes(x=x_predict, y=Ef, ymin=Ef-sigma, ymax=Ef+sigma),
              fill="grey80") + # Var
  geom_line(aes(color=variable)) + #REPLICATES
  geom_line(data=gp,aes(x=x_predict,y=Ef), size=1) + #MEAN
  geom_point(data=obs,aes(x=x,y=y)) +  #OBSERVED DATA
  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
  xlab("input, x")
```

------

# greta

Can we express the same operations in greta?

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(greta)
library(greta.gp)
```

```{r}
# kernel & GP
y <- obs$y
x <- obs$x

# hyperparameters
rbf_var <- 10 # lognormal(0, 1)
rbf_len <- 0.5 #lognormal(0, .01)
obs_sd <- .01 #lognormal(0, .01)

# kernel & GP
kernel <- rbf(rbf_len, rbf_var)
f <- gp(x, kernel)

# likelihood -- not clear how this influences f_plot?
distribution(y) <- normal(f, obs_sd)

# prediction
f_plot <- project(f, x_plot)
```

```{r}
# fit the model by Hamiltonian Monte Carlo
m <- model(f_plot)
draws <- mcmc(m, warmup = 0)
```
```

### 2019-10-26 notebook/greta-gp.Rmd
greta gp exploration

```{r}
# simulate toy data
#x <- runif(20, 0, 10)
#y <- sin(x) + rnorm(20, 0, 0.5)
#x_plot <- seq(-1, 11, length.out = 200)
```

@@

# hyperparameters
rbf_var <- lognormal(0, 1)
rbf_len <- lognormal(0, .01)
obs_sd <- lognormal(0, .01)

# kernel & GP
kernel <- rbf(rbf_len, rbf_var) + bias(1)
f <- gp(x, kernel)

# likelihood
distribution(y) <- normal(f, obs_sd)

@@

```{r}
# fit the model by Hamiltonian Monte Carlo
m <- model(f_plot)
draws <- mcmc(m, one_by_one = TRUE)
```

```{r}
# plot 200 posterior samples
#plot(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 10), ylim = c(-2.5, 2.5))

plot(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 2), ylim = c(-0.1, 0.15))
for (i in 1:200) {
  lines(draws[[1]][i, ] ~ x_plot,
        lwd = 2,
        col = rgb(0.7, 0.1, 0.4, 0.1))  
}
```

>>>>

```{r}
# simulate toy data
n<- 10
x <- runif(n, 0, 10)
y <- sin(x) + rnorm(n, 0, 0.5)
x_plot <- seq(-1, 14, length.out = 200)
```

@@

# hyperparameters
rbf_var <- 10 # lognormal(0, 1)
rbf_len <- 0.5 #lognormal(0, .01)
obs_sd <- .01 #lognormal(0, .01)

# kernel & GP
kernel <- rbf(rbf_len, rbf_var)
f <- gp(x, kernel)

# likelihood -- not clear how this influences f_plot?
distribution(y) <- normal(f, obs_sd)

@@

```{r}
# fit the model by Hamiltonian Monte Carlo
m <- model(f_plot)
draws <- mcmc(m, warmup = 0)
```

```{r}
# plot 200 posterior samples
plot(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 14), ylim = c(-3, 3))

#plot(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 2), ylim = c(-0.1, 0.15))
for (i in 1:200) {
  lines(draws[[1]][i, ] ~ x_plot,
        lwd = 2,
        col = rgb(0.7, 0.1, 0.4, 0.1))  
}
points(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 14), ylim = c(-3, 3))
